Explore the SingleR results

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
reorder nbatlas dotplot

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("barcodes", "sample_id", "library_id", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there are repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode NAs to "unknown_consensus" and remove full consensus labels
  dplyr::mutate(validation_group_annotation = ifelse(
    is.na(validation_group_annotation),
    "unknown_consensus",
    validation_group_annotation
  )) |>
  dplyr::select(-consensus_annotation) |>
  dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id"))

# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  # rename to make annotation sources more clear
  dplyr::rename(
    "consensus_validation" = validation_group_annotation,
    "singler" = pruned.labels
  ) |>
  # pivot longer for wrangling
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type",
    values_to = "label"
  ) |> 
  harmonize_celltypes(label, label_recoded)
```

@@

```{r}
celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler") |>
  dplyr::select(label, label_recoded) |>
  dplyr::distinct() |>
  dplyr::group_by(label_recoded) |>
  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
  dplyr::arrange(label_recoded) |>
  dplyr::rename(
    "Label shown in figures" = label_recoded,
    "Underlying NBAltas label(s)" = all_labels
  ) |>
  knitr::kable()
```

@@

```{r, fig.width = 24, fig.height = 12}
# we want to plot the literal labels here
full_celltype_df <- celltype_df |>
  dplyr::select(
    cell_id, 
    # the dotplot function expects this column name
    label_recoded = pruned.labels
  ) |>
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      # collapse the dendritic cells & monocytes to match what's in the marker genes
      stringr::str_detect(label_recoded, "cDC") ~ "cDC", 
      stringr::str_detect(label_recoded, "monocyte") ~ "Monocyte",
      label_recoded == "NE" ~ "Neuroendocrine",
      .default = label_recoded
  )) |>
  dplyr::filter(label_recoded != "unknown_singler")

# get total number of cells per final annotation group and set up y_label
total_cells_df <- full_celltype_df |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::arrange(desc(total_cells)) |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))

singler_order <- total_cells_df$y_label
total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
nbatlas_bar_order <- total_cells_df$label_recoded
```

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  singler_df = full_celltype_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```
For validation, we expect to see high marker gene expression along the diagonal of the dotplot where cell type labels match with their marker genes.
We do broadly see high expression along the diagonal, which means our labeled cells have some amount of similar signal to those in NBAtlas.
Some sets of marker genes show high expression across multiple cell type categories in particular for closely related cell types (e.g. monocyte, macrophage, myeloid, dendritic, or T-cells and NK-cells), but there do not appear to be any strongly unexpected gene expression patterns.

>>>>

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("barcodes", "sample_id", "library_id", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there are repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode NAs to "unknown_consensus" and remove full consensus labels
  dplyr::mutate(validation_group_annotation = ifelse(
    is.na(validation_group_annotation),
    "unknown_consensus",
    validation_group_annotation
  )) |>
  dplyr::select(-consensus_annotation) |>
  dplyr::left_join(singler_df, by = c("cell_id", "barcodes", "library_id")) |>
  # we do specifically need to change pruned.labels from NE -> Neuroendocrine
  dplyr::mutate(
    pruned.labels = ifelse(pruned.labels == "NE", "Neuroendocrine", pruned.labels)
  )

# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  # rename to make annotation sources more clear
  dplyr::rename(
    "consensus_validation" = validation_group_annotation,
    "singler" = pruned.labels
  ) |>
  # pivot longer for wrangling
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type",
    values_to = "label"
  ) |> 
  harmonize_celltypes(label, label_recoded)

@@

```{r}
label_map_df <- celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler", label != "unknown_singler") |>
  dplyr::select(label, label_recoded) |>
  dplyr::distinct()

label_map_df |>
  dplyr::group_by(label_recoded) |>
  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
  dplyr::arrange(label_recoded) |>
  dplyr::rename(
    "Label shown in figures" = label_recoded,
    "Underlying NBAltas label(s)" = all_labels
  ) |>
  knitr::kable()
```

@@

```{r, fig.width = 24, fig.height = 12}
# we want to plot the literal labels here
full_celltype_df <- celltype_df |>
  dplyr::filter(pruned.labels != "unknown_singler") |>
  dplyr::select(cell_id, label = pruned.labels) |>
  # collapse certain labels that are grouped in the marker genes
  dplyr::mutate(
   label = dplyr::case_when(
     stringr::str_detect(label, "cDC") ~ "cDC",
     stringr::str_detect(label, "monocyte") ~ "Monocyte",
     .default = label
  ))

# Also update label_map_df to collapse those labels
label_map_df <- label_map_df |>
  dplyr::mutate(
   label = dplyr::case_when(
     stringr::str_detect(label, "cDC") ~ "cDC",
     stringr::str_detect(label, "monocyte") ~ "Monocyte",
     .default = label)
  ) |>
  unique()

# Prepare an order of label_recoded to help set the order for the literal underlying labels
label_coded_factor <- celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler", label_recoded != "unknown_singler") |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::arrange(desc(total_cells)) |>
  dplyr::pull(label_recoded)
label_recoded_factor <- factor(label_coded_factor, levels = label_coded_factor)

# get total number of cells per final annotation group and set up y_label
total_cells_df <- full_celltype_df |>
  dplyr::count(label, name = "total_cells") |>
  # order so cell types are grouped
  dplyr::inner_join(label_map_df, by = "label" ) |>
  dplyr::mutate(
    label_recoded = as.factor(label_recoded),
    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
  ) |>
  dplyr::arrange(label_recoded) |>
  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
  # rename as expected for dotplot function - label_recoded needs to be the main column
  dplyr::select(-label_recoded) |>
  dplyr::rename(label_recoded = label)

total_cells_df$y_label <- factor(total_cells_df$y_label, levels =  total_cells_df$y_label)
nbatlas_bar_order <- total_cells_df$label_recoded
```

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  # rename this after we've done all the wrangling
  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```
For validation, we expect to see high marker gene expression along the diagonal of the dotplot where cell type labels match with their marker genes.
We do broadly see high expression along the diagonal, which means our labeled cells have some amount of similar signal to those in NBAtlas.
Some sets of marker genes show high expression across multiple cell type categories in particular for closely related cell types (e.g. monocyte, macrophage, myeloid, dendritic, or T-cells and NK-cells), but there do not appear to be any strongly unexpected gene expression patterns.
